---
title: "CjsModel: Cormack-Jolly-SeberSCR Model Object"
author: "Richard Glennie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CjsModel: Cormack-Jolly-Seber SCR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE, 
  collapse = TRUE,
  fig.width = 7, 
  fig.height = 7, 
  comment = "#>"
)
```

In this vignette, I describe how Cormack-Jolly-Seber (CJS) spatial capture-recapture (SCR) models can be used within an <code>R6</code> object of the <code>CjsModel</code> class. I recommend you read
the vignettes on the ScrData and ScrModel classes before reading this one. The folder inst/examples/ in the [repository](https://github.com/r-glennie/openpopscr) contains example code.

```{r load_pkg, eval = TRUE}
library(openpopscr)
library(secr)
```
```{r secretload, eval = TRUE, include = FALSE}
data("cjsmodel_example")
```
## Simulating data 

The <code>openpopscr</code> package uses the <code>ScrData</code> class to store data. See the vignette on the <code>ScrData</code> class to see how to create the data object from your own raw data. In this section, I simulate a CJS SCR survey. The simulation function is useful when you want to try out a new method where you know what the correct parameter estimates are, you want to test how robust a method is to violations of its assumptions, or you want to see how data simulated from a model compare with real data.

Cormack-Jolly-Seber models are used to describe the pattern of detections for 
individuals that are known to be alive from some occasion onwards. That is, one 
conditions on the first detection of the individual and estimates survival and 
detectability from the repeated detections after that. No inference is extended
to the entire population other than the assumption that the survival probability
for captured individuals is the same as for the population as a whole. 

To simulate a survey, I require true detection and survival parameters, a detector layout, and a mesh.  

```{r setuptruepar}
# set true parameters 
true_par <- list(lambda0 = 0.5, sigma = 20, phi = 0.7)

# set abundance 
N <- 100

# make detectors array 
detectors <- make.grid(nx = 7, ny = 7, spacing = 20, detector = "count")

# make mesh 
mesh <- make.mask(detectors, buffer = 100, nx = 64, ny = 64, type = "trapbuffer")

# set number of occasions to simulate
n_occasions <- 5
```

To simulate the data, I set a random seed so that the results will be the same every time the code is run. This is so that you, the reader, can re-run the code and produce the same output as me. In practice, you may not want to set the seed if you want the output from the simulation to change randomly. 

```{r simulatescr}
scrdat <- simulate_cjs_openscr(true_par, N, n_occasions, detectors, mesh, seed = 19539)
```

You will notice the <code>simulate_cjs_openscr</code> function produces useful output to let the user know what is going on. If you want to switch this off, for example if you are running lots of simulations and do not want this output being produced over and over, then you can set <code>print = FALSE</code> just as I have set the seed above. 

In the simulated survey, `r scrdat$n()` unique individual were detected at least once. More summary information can be found by typing the name of the object into the console: 

```{r look_at_data, eval = TRUE}
scrdat
```

 
## Fitting a model 

To create a CjsModel, you need to specify a formula for each detection parameter
and the survival parameter; also, you need starting values for all parameters.
It is best to choose starting values for the parameters that are biologically
reasonable and match your data. You could simulate some data using proposed
starting values as true parameters to see if they lead to similar data. Poor
starting values can lead to spurious convergence in complicated models, models
with many parameters. The function <code>get_start_value</code> can be used to
produce a reasonable guess for each parameter.

```{r setupscrmod}

# set each parameter to a be constant 
form <- list(lambda0 ~ 1, 
             sigma  ~ 1,
             phi ~ 1)

# get some starting values based on data 
start <- get_start_values(scrdat, model = "CjsModel")

# check starting values are reasonable, they seem ok 
start 
```

Before fitting the model, I create the model object: 

```{r modelobj}
mod <- CjsModel$new(form, scrdat, start)
```

Once the model object has been created, you can fit the model using the fit
function:
```{r fitmodel}
mod$fit()
```

If you have the output printed, the "checking convergence" message reports
whether convergence has been reached and if it has not it reports the code
output by <code>optim</code> that tells you why it failed to converge. 

# Reading model output 

Once a model has been fit, you can type its name into the console and some
summary results are printed: 
```{r printres, eval = TRUE}
mod
```

The parameters are reported on the link scale; this is because when covariates
are involved, it is easier to interpret their effects. The package <code>openpopscr</code>
uses the log-link function for all of these parameters. So, you would take
the exponential of these reported numbers to obtain the parameter values on the
response scale. 

Alternatively, you can use the <code>get_par</code> function: 
```{r getpar, eval = TRUE}
mod$get_par("lambda0", k = 1, j = 1)
mod$get_par("sigma", k = 1, j = 1)
mod$get_par("phi", k = 1)
```

I specify $k = 1, j = 1$ in the functions above for $\sigma, \lambda_0$ because by default
<code>get_par</code> returns the value of these parameters for every occasion and detector. This is
useful when they vary by occasion, but here they are constants, so I only want
their value on a single occasion, here I chose the first occasion. 

If you want to access the reported results, they are stored as a list which is
returned by the function <code>estimates</code>:
```{r getest, eval = TRUE}
mod$estimates()
```

The model AIC can be computed in the usual way: 
```{r getaic, eval = FALSE}
AIC(mod)
```

## Covariates 

The detection parameters can change with observed covariate values. Covariates
can change by detector or by occasion. At present, individual-level
covariates are not implemented in <code>openpopscr</code>. By default, the
covariate $t$ is created to allow for parameters to depend on occasion. 

As an example, one can fit a model where survival changes depending on 
anthropogenic distriburance (measured by some continuous covariate): 
```{r survt}
# simulate a disturbance covariate (% deforestation for example)
set.seed(385638)
disturb <- runif(scrdat$n_occasions()) 
# add covariate 
scrdat$add_covariate("disturb", disturb, "k")
# copy formulae from previous model
form_surv <- form
# update formula for phi to include age 
form_surv[[3]]<- phi ~ disturb 
```
```{r survmake}
# create new model object 
mod_surv <- CjsModel$new(form_surv, scrdat, start)
# fit model
mod_surv$fit()
```
```{r survanal, eval = TRUE}
# look at results
mod_surv
# predict phi for each occasion
mod_surv$get_par("phi")
```

Covariates that change over time, by detector, and by detector-time can be
included in the same way. See ScrData and ScrModel vignettes for more details on
loading covariates into the ScrData object and fitting models with covariates.

## Time and Robust Design

There are two further arguments you can give when creating a ScrData object: 
time and primary. 

The time option requires a vector that contains the start time of 
each occasion. The time unit is the user's choice. For example, if occasions
occur in Weeks 1, 3, 4 of a particular month, this could be specified in a time
vector as <code>c(0, 3, 4)</code> (in weeks) or <code>c(0, 14, 21)</code> (in days). 
```{r timescrdata, eval = FALSE}
scrdat <- ScrData$new(capthist, mesh, time = time)
```
Once time is specified, the survival estimates will be the estimated probability
of survival in one time unit, so it will depend on what time unit you used. 

The primary argument relates to robust design surveys. In a robust design, 
there are both primary and secondary occasions for sampling.
Secondary occasions take place close together in time and it is assumed the population 
is closed between them. Population change occurs between primary occasions. 
The survival parameter between primary occasions is of interest. 

For <code>ScrData</code>, you supply a single capthist object (combined across all primary occasions)
and then specify in the <code>primary</code> argument which primary each secondary occasion belongs to. 
For example, suppose a survey has 3 primary occasions with 2, 4, 4 secondary occasions respectively. The
 capthist object would then have 10 occasions in it and the primary vector would be <code>c(1,1,2,2,2,2,3,3,3,3)</code>
```{r robustdesign, eval = FALSE}
scrdat <- ScrData$new(capthist, mesh, primary = primary)
```
An example of a robust design analysis is included inst/examples/cjs_robust_example.R

## Transience 

For Cjs models, transience can play an important role. When individuals are transient,
their activity can pass in and out of range of detection; individuals that permanently
pass out of range of detection are best explained, by the CjsModel, as having died
and so survival probability is underestimated. This forces one to estimate only
apparent survival, not true survival. When transience is accounted for, one can
more defendably state that they have estimated true survival. 

You can simulate Cjs SCR surveys with transience: 
```{r cjstrans}
# set truth 
true_par <- list(lambda0 = 1, sigma = 20, phi = 0.7, sd = 10)
# make detectors array 
detectors <- make.grid(nx = 7, ny = 7, spacing = 20, detector = "count")
# make mesh 
mesh <- make.mask(detectors, buffer = 100, nx = 64, ny = 64, type = "trapbuffer")
# set number of occasions to simulate
n_occasions <- 5
# set number of individuals tracked
N <- 100
# simulate ScrData 
scrtransdat <- simulate_cjs_openscr(true_par, 
                                    N, 
                                    n_occasions, 
                                    detectors, 
                                    mesh, 
                                    move = TRUE, 
                                    seed = 95811)
```

I will first fit a stationary CJSModel to these data: 
```{r statscr}
# set formulae and start values 
stat_form <- list(lambda0 ~ 1, sigma ~ 1, phi ~ 1)
start <- get_start_values(scrtransdat, model = "CjsModel")
```
```{r statmake, include=FALSE}
# create model object 
stat <- CjsModel$new(stat_form, scrtransdat, start)
# fit model 
stat$fit()
```
```{r statanal, eval = TRUE}
# look at results 
stat
# look at parameters on response scale 
stat$get_par("lambda0", k = 1, j = 1)
stat$get_par("sigma", k = 1, j = 1)
stat$get_par("phi", k = 1)
```

Given we know the truth, the estimates are a little off: $\lambda_0$ is far too 
low, $\sigma$ is too high, and $\phi$ is a litte low.   

I now fit the transient model. The transient model class is called
<code>CjsTransientModel</code>. We must also now specify a start value for the
movement parameter.

```{r fittransmod}
# specify formulae for parameters 
form <- list(lambda0 ~ 1, 
             sigma  ~ 1,
             phi ~ 1,
             sd ~ 1)

start <- get_start_values(scrtransdat, model = "CjsTransientModel")
```
```{r fittedtransmake}
trans <- CjsTransientModel$new(form, scrtransdat, start)
```

I now fit the model as usual:
```{r fittrans}
# fit model 
trans$fit()
```
```{r transanal, eval = TRUE}
# look at results 
trans
# look at parameters on response scale 
trans$get_par("lambda0", k = 1, j = 1)
trans$get_par("sigma", k = 1, j = 1)
trans$get_par("sd", k = 1)
trans$get_par("phi", k = 1)
```

The survival probability is marginally higher and the estimate of activity range 
reduced by accounting for transience. 

The two models can be compared by AIC. The transient model is preferred. 
