---
title: "openpopscr"
author: "R. Glennie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{openpopscr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

During capture-recapture surveys that span a time comparable to the 
temporal scale at which a population evolves, animals can die, migrate, or be born. 
Open population capture-recapture models allow for these population processes. 

This package can fit open population Jolly-Seber spatial capture-recapture models by 
maximum likelihood. This guide describes how to read in your data, fit some 
models, and obtain inference.


## Reading in Data 

It is assumed you have a `capthist` and `mesh` object for your data, built using 
the `secr` package. Here, example data provided with this package is 
used.  

```{r, fig.show='hold'}
# load package 
library(openpopscr)
# load example data 
data("openpop_example_data")
capture_history <- openpop_example_data$capture.history
mesh <- openpop_example_data$mesh
```

The example survey consists a 7 x 7 array of detectors with 10 capture occasions.
Over the survey, `r dim(capture_history)[1] ` unique individuals are detected. 
Detectors record the number of times each individual was seen within each 
occasion and individuals can be seen on multiple detectors per occasion.

In `openpopscr`, the data is stored in a single object called an ScrData object.
A new ScrData object is created using the $new method with the capture history
and mesh provided as arguments. Typing the name of the ScrData object into 
the console returns the visual and numerical summary of the data provided 
by the `secr` package.

```{r, fig.show='hold', fig.align='center', fig.width = 5, fig.height = 5}
scr_dat <- ScrData$new(capture_history, mesh)
scr_dat
```

If the capture occasions occur at irregular intervals, this should be included. 
A vector of times at which each occasion began, `t`, can be used to create the data 
object with `ScrData$new(capture_history, mesh, t)`. 

## Data exploration 

The ScrData object can be queried to return information about the data stored. 
The commands ``scr_dat$capthist()``, ``scr_dat$traps()``, and ``scr_dat$mesh()``
return the capture history, traps, and mesh objects respectively. 

Further information on the data is available:
```{r}
# get number of unique individuals seen 
scr_dat$n()
```
```{r}
# get number of occasions in the survey 
scr_dat$n_occasions()
```
```{r}
# get number of traps in the survey 
scr_dat$n_traps()
```
```{r}
# get number of mesh points in the survey 
scr_dat$n_meshpts()
```
```{r}
# get total area of the mesh 
scr_dat$area()
```
```{r}
# get matrix of distances between each trap and each mesh point 
dist <- scr_dat$distances()
# e.g. distances from trap 1 to first 6 mesh points 
head(dist[1,])
```

## Fitting the Jolly-Seber model

The Jolly-Seber model is another type of object, `JsModel`. It is also created
with the `$new` command with arguments giving the ScrData object and initial
parameter values.

Parameters can depend on covariates that vary by detector or by occasion. 
Parameters cannot depend on individual-level or spatial covariates. Covariates
are included similar to formulae used to specify a linear model in `R` using the
`lm` command; however, as there are multiple parameters that may depend on 
covariates, the formulae are given in a list. 

For a model where all parameters are constant, an intercept-only formula is 
specified for each parameter. Sensible starting values for each parameter 
are also required: for beta, phi and D this ought to be possible to deduce from
knowledge of the study species; reasonable values for lambda0 and sigma can be 
found by fitting a standard secr model to the data. 
```{r}
# list formulae for parameters (intercept-only models)
par <- list(lambda0 ~ 1, 
            sigma ~ 1, 
            beta ~ 1, 
            phi ~ 1)

# guess a reasonable for value for each parameter 
start <- list(lambda0 = 2.0, 
              sigma = 100, 
              beta = 0.2, 
              phi = 0.5, 
              D = 100)

# create JsModel object 
mod <- JsModel$new(par, scr_dat, start)
```

Creating the model object does not fit the model, the `$fit` command does this.
```{r, eval = FALSE}
mod$fit()
```

While fitting, the likelihood will be printed for each evaluation. Once fit, 
parameter estimates can be viewed by typing the name of the model object
into the console. 
```{r, results="hide"}
mod <- readRDS("fitted_mod_k.Rds")
```
```{r}
mod
```

The parameter estimates table gives estimates on the link scale. This is useful
later when covariates are used. The link functions for each parameter are set: 
lambda0, sigma, and D have log-link functions, beta has a multilogistic link 
function, and phi a logit link function.  


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
