---
title: "ScrData: Spatial Capture-Recapture Data Object"
author: "Richard Glennie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7, 
  fig.height = 7, 
  comment = "#>"
)
```

The ScrData class is an [<code>R6</code> class](https://cran.r-project.org/web/packages/R6/vignettes/Introduction.html). <code>R6 classes</code> are used to create objects. Objects can have properties and functions. These objects are convenient because they can be stored and used as a single <code>R</code> object, but contain within them all the information and methods related to them. 

In this vignette, I describe how spatial capture-recapture data can be stored within an <code>R6</code> object of the <code>ScrData</code> class. 

```{r load_pkg}
library(openpopscr)
library(secr)
```

## Raw data 

The [<code>secr</code> package](https://cran.r-project.org/package=secr) contains the functions <code>read.traps</code>, <code>make.capthist</code>, and <code>make.mask</code>. They can be used to convert raw data often collected in SCR surveys into a capthist object and mask object. These functions are available in the <code>openpopscr</code> package also. 

On a side note, I call the mask object a mesh. The terms are interchangable. 

For this vignette, I will use the <code>stoatDNA</code> data available from the <code>secr</code> package. These data
come from a DNA hair trapping survey on stoats. 

```{r load_data}
# load data 
data("stoat")
captures <- stoat$captures
traps <- stoat$traps

# look at capture history structure 
str(captures)
# look at detector locations 
str(traps)
```

The <code>read.traps</code> function will take a data frame with x, y coordinates for each detector and create
 the traps object. The detector type here is "proximity" because DNA hair traps only record whether the stoat 
 was detected or not by the trap, not how many times it was detected (that would be a "count" type detector). Also, with proximity detectors it is possible for a single individual to be
 detected on more than one trap during a single occasion; if this is not the case,
 then the traps are called "multi" type.  
 
```{r make_traps}
 detectors <- read.traps(data = traps, detector = "proximity")
```

The <code>make.capthist</code> function creates the capthist object. 

```{r make_capthist}
capthist <- make.capthist(captures, detectors)
```


This data does not have a custom mesh, so I create one using <code>make.mask</code>: 
```{r make_mask}
mesh <- make.mask(detectors)
```

## A simple ScrData object 

The ScrData object requires at least the capthist and mesh objects. R6 objects are created using the <code>$new</code> function. 

```{r simple_scrdata}
scrdat <- ScrData$new(capthist, mesh)
```

Typing the name of the object into the terminal will create a plot and capthist summary information from the secr package. In the plot, 
detectors are red crosses, the mesh points are grey, and captures are coloured by individual. 

```{r summmary}
scrdat
```

The functions of the ScrData object can be used to extract the capture history, mesh, or other information: 
```{r data_summary}
# get number of unique individuals seen 
scrdat$n()

# get number of occasions in the survey 
scrdat$n_occasions()

# get number of traps in the survey 
scrdat$n_traps()

# get number of mesh points in the survey 
scrdat$n_meshpts()

# get total area of the mesh in square kilometres 
scrdat$area()

# get matrix of distances between each trap and each mesh point 
dist <- scrdat$distances()
# e.g. distances from trap 1 to first 6 mesh points 
head(dist[1,])
```

One thing to note is the <code>area</code> function, it returns the area in square kilometres, not hectares, the standard unit of area for the <code>secr</code> package. 

## Time 

Occasions may occur at irregular intervals. This information can be included in the ScrData object when it is created. For example, suppose the data were collected on $3$ days, then a week later, for $4$ days. The time index of each occasion in days would then be 
```{r time}
time <- c(1:3, 10:13)
```

This is included in the ScrData object when <code>$new</code> is called. 

```{r timeincluded, eval = FALSE}
scrdat <- ScrData$new(capthist, mesh, time = time)
```

It can be accessed using the command 
```{r gettime, eval = FALSE}
scrdat$time()
```

When time is not specified, it is assumed that occasions occurred in regular intervals of 
duration $1$. 

## Covariates 

The ScrData object can handle three kinds of covariates: detector-specific, temporal, or both. 

Detector-specific covariates are called $j$ type covariates, because $j$ is typically the mathematical index used for detectors. 
Suppose there is a variable $x$ that could affect how detectors operate. I create a imaginary variable here: 

```{r x_cov}
x <- factor(sample(c("bad", "good"), size = nrow(detectors), replace = TRUE))
```

This is a factor variable that labels each detector as a "good" or "bad" detector. 

The covariate can be included when creating the ScrData object: 

```{r cov_make_scrdata, eval = FALSE}
scrdat <- ScrData$new(capthist, mesh, cov = list(x = x), cov_type = c("j"))
```

The <code>cov</code> argument must be a list with each covariate named. The <code>cov_type</code> argument specifies the type of each covariate; it must be ordered in the same way as the <code>cov</code> argument. 

Alternatively, the covariate can be added to the ScrData object after it has been created using the <code>add_covariate</code> function:

```{r add_cov}
scrdat$add_covariate("x", x, "j")
```

The <code>add_covariate</code> function requires the covariate name, the covariate data frame itself, and the covariate type. 

Temporal covariates are added in the same way, except the <code>cov_type</code> is "k", the index commonly used to denote occasion. 

Once covariates have been included, you can extract the covariates for any particular detector or time using the <code>covs</code> function. To list all covariates, just call the function with no arguments: 

```{covs}
scrdat$covs()
```

Notice, that the covariate $t$ is automatically created; it is the index of each occasion. 

Covariates can be subset by detector or time: 
```{r subsetcov}
# get covariates for all detectors on occasion 2
scrdat$covs(k = 2)
# get covariates for detector 7 on all occasions 
scrdat$covs(j = 7)
# get covariates for detector 5 on occasion 3 
scrdat$covs(j = 5, k = 3)
# get covariates for detectors 1 to 10 on occasions 3 and 5
scrdat$covs(j = 1:10, k = c(3, 5))
```

