---
title: "StrataModel: Spatial Capture-Recapture with Strata"
author: "Richard Glennie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7, 
  fig.height = 7, 
  comment = "#>"
)
```

The StrataModel class is an [<code>R6</code> class](https://cran.r-project.org/web/packages/R6/vignettes/Introduction.html). <code>R6 classes</code> are used to create objects. Objects can have properties and functions. These objects are convenient because they can be stored and used as a single <code>R</code> object, but contain within them all the information and methods related to them. 

In this vignette, I describe how other types of model classes (ScrModel, CjsModel, JsTransientModel) can be 
fit to multiple strata using the StataModel class.

```{r load_pkg}
library(openpopscr)
library(secr)
```
```{r secretload, include = FALSE}
data("stratamodel_example")
obj <- stratamodel_example$obj
scrdat1 <- stratamodel_example$scrdat1
scrdat2 <- stratamodel_example$scrdat2
scrdat3 <- stratamodel_example$scrdat3
scrdat <- stratamodel_example$scrdat
```
## Simulate data

I will simulate a survey that is conducted over three strata. The idea is strata
may share some parameters (i.e., parameters are equal among strata) and for
other parameters, each stratum will have its own particular value. Strata are
treated independently conditional on the parameters. 

Here, all strata have the same value of lambda0, but different sigma and D parameters. 
I create a <code>ScrData</code> object for each stratum seperately and then
store these in a list. I use the same detector array setup and mesh in each stratum, but
these could differ. 

```{r simdata, eval = FALSE}
true_par <- list(s1 = list(D = 1000, lambda0 = 2, sigma = 20), 
                 s2= list(D = 500, lambda0 = 2, sigma = 20), 
                 s3 = list(D = 1000, lambda0 = 2, sigma = 10))

# make detectors array 
detectors <- make.grid(nx = 7, ny = 7, spacing = 20, detector = "multi")

# make mesh 
mesh <- make.mask(detectors, buffer = 100, nx = 64, ny = 64, type = "trapbuffer")

# set number of occasions to simulate
n_occasions <- 5 

# simulate for each stratum
scrdat1 <- simulate_scr(true_par[[1]], n_occasions, detectors, mesh, seed = 15483)
scrdat2 <- simulate_scr(true_par[[2]], n_occasions, detectors, mesh, seed = 79696)
scrdat3 <- simulate_scr(true_par[[3]], n_occasions, detectors, mesh, seed = 43523)

scrdat <- list(s1 = scrdat1, s2 = scrdat2, s3 = scrdat3)
```

```{r seedata}
scrdat1
scrdat2
scrdat3

```

## Fitting the model

The <code>StrataModel</code> class works by specifying a formula for each 
parameter that is to be common among all strata, then separately specifying a
formula for each stratum for each parameter that is to be particular to strata. 
This allows stratum-specific parameters to depend on different covariate combinations. 

When creating the <code>StrataModel</code> you must also specify what type of 
model you wish to fit to each stratum. Here, I fit the <code>ScrModel</code>.
You can also specify any other model class available within the <code>openpopscr</code>
package. 

```{r createobj, eval = FALSE}
shared_form <- list(lambda0 ~ 1)

private_form <- list(s1 = list(sigma ~ 1),
                     s2 = list(sigma ~ 1), 
                     s3 = list(sigma ~ 1))


start <- list(lambda0 = 2, sigma = 20, D = 1000)

obj <- StrataModel$new(scrdat, "ScrModel", shared_form, private_form, start)
```

```{r fitmodel, eval = FALSE}
obj$fit()
```

Once fit, the object's name can be typed into the terminal to return 
the results for each stratum. 

```{r results}
obj
```

Furthermore, the <code>get_object</code> function can be used to extract
the model object that corresponds to each stratum. You can then use the
functions that model object has to extract estimates or other quantities. 

```{r getllk}
# e.g. get likelihood value at maximum for stratum 1 model 
obj$get_object(1)$mle_llk()
```

You can extract parameters for each stratum using the <code>get_par</code>
function:

```{r object}
# should be the same for all strata
obj$get_par("lambda0", k = 1)
# differ between strata (in simulation strata 1 & 2 have same, 3 differs)
obj$get_par("sigma", k = 1)
# density by stratum
obj$get_par("D")
```
